---
title: "A Unified Probabilistic Framework for Non-Stationary Heart Rate Variability Analysis"
subtitle: "Modeling the Dynamic Evolution of Autonomic Control"
date-modified: today
bibliography: "bibliography.bib"
execute: 
  echo: false
  message: false
  warning: false
  error: false
author: 
  - name: "Matías Castillo-Aguilar"
    email: matias.castillo@umag.cl
    orcid: 0000-0001-7291-247X
  - name: "Cristian Núñez-Espinosa"
    email: cristian.nunez@umag.cl
    orcid: 0000-0002-9896-7062
format: 
  html: 
    fig-format: svg
    title-block-banner: true
    title-block-style: manuscript
    toc: true
    toc-location: right
    toc-depth: 2
    comments: 
      hypothesis: 
        theme: clean
other-links: 
  - text: "Stan Model"
    href: "models/rri_model.stan"
    icon: file-code
fig-width: 8
fig-height: 6
editor_options: 
  chunk_output_type: console
---

# Introduction

A central goal in contemporary biomedical science is to mechanistically link the dynamics of physiological systems to human health and disease. A primary obstacle to this goal is the inferential gap between what we can measure with high-temporal resolution via wearable sensors, and the unobservable neurobiological processes that generate these measurements [@huber2025brain]. This "inverse problem" is nowhere more apparent than in the study of the autonomic nervous system (ANS), where heart rate variability (HRV) is the dominant non-invasive proxy [@shaffer2017overview]. The R-R interval (RRi) time-series is an invaluable signal, yet it is also a profound analytical challenge: it is a complex, integrated output reflecting the ceaseless, dynamic interplay of multiple regulatory systems [@arakaki2023connection]. Therefore, progress is contingent upon solving a fundamental problem of principled inverse inference: the development of a formal methodology to deconstruct this composite signal into quantitative, interpretable, and falsifiable estimates of its underlying generative processes. The urgency of this problem is underscored by its direct relevance to major public health burdens, as autonomic dysregulation is a core feature of disorders ranging from cardiovascular disease to anxiety and post-traumatic stress [@cheng2022heart; @tomasi2024heart; @schneider2020autonomic], where a precise characterization of system dynamics is essential for diagnosis and treatment [@orini2023long].

The analytical tools currently employed by the majority of researchers are fundamentally unsuited for this task. The prevailing paradigm, which relies on calculating summary statistics within fixed temporal windows, imposes a static, linear worldview on a system that is intrinsically dynamic and non-linear. This approach is predicated on an assumption of stationarity that is inherently violated during any physiological transition of scientific or clinical interest [@shaffer2020critical; @kim2021ultra]. This is not a minor statistical infraction; it is a fundamental limitation for mechanistic inference that produces misleading and often uninterpretable results [@kim2021ultra]. For instance, by averaging across a period of stress recovery, such methods conflate the initial sympathetic response with the subsequent parasympathetic rebound, yielding a single, opaque number that masks the system's adaptive capacity [@shaffer2020critical; @kim2021ultra]. This can render it impossible to distinguish a resilient individual who exhibits a rapid and robust recovery from a vulnerable one whose recovery is sluggish, a critical biomarker of pathology [@lu2023uncertainties]. This "discretize-and-summarize" approach systematically discards the temporal structure of the data, thereby precluding the direct estimation of core mechanistic properties such as response latencies, rates of change, and adaptive magnitudes, which constitute the very essence of the physiological process [@tiwari2021analysis].

Acknowledging these limitations, more sophisticated signal processing and modeling techniques have been proposed, yet these too fail to resolve the core inferential challenge. They present a false choice between phenomenological description and non-interpretable abstraction. Time-frequency decompositions, such as the wavelet transform, excel at providing detailed visualizations of spectral dynamics, making them useful for exploratory analysis [@german2014wavelet]. However, they are fundamentally descriptive; they do not instantiate a formal, generative model of the underlying biology. They can show that a change occurred, but they cannot provide a parameterized, testable estimate of what changed [@guo2022review]. Conversely, advanced modeling traditions like state-space models offer mathematical sophistication for tracking and forecasting a system's evolution [@martin2018improving; @rosas2024bayesian]. While powerful, their typical formulations trade physiological fidelity for mathematical convenience; the latent states they infer are defined by statistical properties rather than biological identity, lacking a direct, one-to-one mapping to constructs like vagal tone or sympathetic drive [@rosas2024bayesian; @liu2024efficient]. Consequently, the field remains methodologically stalled, caught between descriptive tools that cannot test mechanisms and abstract models whose mechanisms are not directly interpretable.

We argue that bridging the gap between data and mechanism requires a fundamental evolution in analytical philosophy: a shift from model-agnostic data description to the construction of hypothesis-driven, generative models. This approach moves beyond simply characterizing what the data looks like and instead focuses on creating explicit, testable, and quantitative models of the biological machinery that produced it. Here, we introduce and validate such a framework, which recasts physiological analysis as a form of formal hypothesis testing. We posit that a stereotyped autonomic response to an acute perturbation is not random noise but a structured, dynamic process. Consequently, we propose it can be effectively modeled as a continuous-time stochastic process, allowing us to separate the structured response from inherent biological variability. As a specific, falsifiable instantiation of this philosophy, we model the response trajectory using a double-logistic function. This mathematical form is chosen not as a universal truth, but as a parsimonious hypothesis for biphasic dynamics, designed to capture the characteristic onset and subsequent recovery phases [@castillo2025enhancing]. The power of this approach lies in its ability to translate the raw signal into a small set of interpretable mechanistic parameters: the latency, rate, and magnitude of each distinct response phase [@castillo2025enhancing]. This provides a direct, quantitative window into the system's underlying reactivity and resilience.

We implement this approach as a novel generative model within a Bayesian framework. This provides robust parameter estimation with full uncertainty quantification, enables hierarchical modeling of individual and group-level effects, and offers a principled basis for model comparison. This model provides a specific and falsifiable instantiation of our framework, decomposing the RRi series into a time-varying mean, total variance, and a dynamically shifting oscillatory component, all governed by a double-logistic latent dynamic. This paper tests the hypothesis that this model provides a more accurate and useful characterization of autonomic dynamics than existing methods. Through simulation, we demonstrate its capacity to recover known ground-truth parameters, establishing its core validity. We contend that this approach provides the necessary quantitative indices to develop and test mechanistic physiological hypotheses, representing a critical step towards translating complex time-series into falsifiable knowledge.

```{r}
#| include: false
library(data.table)
library(gt)
```

# Results

This section evaluates the proposed quantitative framework. The results are presented in three parts. We first test the model's internal consistency by assessing its ability to recover its own parameters from simulated data. We then benchmark its descriptive fidelity against conventional methods to quantify its effectiveness in characterizing dynamic signals. Finally, we use the framework to generate a structured, phenomenological description of an empirical exercise-response signal, illustrating its potential for generating precise and data-driven hypotheses.

## Model validation and parameter recovery on synthetic data

To establish the model's internal consistency and the identifiability of its parameters, we first assessed its ability to perform inference on synthetic data generated from its own structure. This in-silico procedure serves as a necessary check of the model's self-consistency and the estimation algorithm's ability to navigate the complex posterior geometry. We designed three distinct scenarios to test performance: (A) a canonical response with complete recovery; (B) an incomplete recovery with persistence in both time and frequency domains; and (C) a high-noise condition with a low structured variance fraction ($w$ = 0.6). The ground-truth dynamics are shown in Figure 1, with generative parameter values listed in Table 1. It is critical to note that this procedure tests for inferential integrity under the assumption that the model is correctly specified; it does not test for robustness to model misspecification, a challenge inherent to all empirical data.

![](figures/fig-generated-data.svg){width=100%}
**Figure 1. Ground-truth dynamics of three synthetic scenarios.** The scenarios were designed to test model performance under diverse conditions: (A) a classic sympatho-vagal response, (B) an incomplete recovery with persistence in both time and frequency domains, and (C) a high-noise condition with dissonant (complete time-domain, incomplete spectral) recovery. Panels show the generated RRi data with the true mean signal, the underlying time-domain trajectories for baseline RRi and total SDNN, and the time-varying spectral proportions.

```{r}
#| column: screen-inset-left
readRDS(file = "tables/tbl-1.RDS")
```

**Table 1. Ground-truth parameter values for simulation scenarios.** Ground-truth parameter values used in the generation of three simulation scenarios. The scenarios correspond to (A) a classic sympatho-vagal response, (B) an incomplete recovery with time-domain and spectral persistence, and (C) an incomplete spectral but complete time-domain recovery with high noise.

When fit to this synthetic data, the model demonstrated a high-fidelity reconstruction of the generative process across all three scenarios. Figure 2 presents a visual summary of these results, plotting the model's posterior estimates against the known ground-truth trajectories. The posterior means for the baseline RRi, total SDNN, and the time-varying spectral proportions closely track the true underlying dynamics. Furthermore, the 95% credible intervals, which represent the range of plausible parameter values given the data, consistently envelop the ground-truth trajectories. This latter observation is of particular importance, as it indicates a well-calibrated quantification of uncertainty, a primary objective of the Bayesian approach. The model does not merely provide a point estimate but also a principled measure of its own confidence in that estimate, a feature that is essential for cautious scientific interpretation.

![](figures/fig-model-method.svg){width=100%}

**Figure 2. High-fidelity reconstruction of synthetic data by the generative model.** The model's posterior mean estimates (solid lines) and 95% credible intervals (shaded ribbons) are overlaid on the known ground-truth dynamics (dashed lines) for all three scenarios (A, B, C). The model accurately recovers the continuous trajectories for baseline RRi, total SDNN, and spectral proportions across all conditions.

Beyond recovering the continuous dynamic trajectories, the model also recovered the discrete parameters that govern the shape and timing of these dynamics. Table 2 presents the posterior summaries for these key parameters, comparing the posterior median and 95% credible intervals against their true values. For all parameters across all scenarios, the 95% credible intervals successfully contained the ground-truth values. This result confirms that the model's parameters are practically identifiable from data of this type, at least when the data-generating process aligns with the model's structural assumptions. This successful recovery provides compelling evidence that the parameters are not hopelessly confounded and that the likelihood surface is sufficiently well-behaved to permit robust inference, providing the necessary foundation for the model's application to more complex, real-world data.

```{r}
#| column: screen-inset-left
readRDS(file = "tables/tbl-2.RDS")
```

**Table 2. Posterior recovery of ground-truth parameters.** This table compares the known ground-truth parameter values to the model's posterior estimates (median and 95% credible interval). The model successfully recovers all parameters, with the true values consistently falling within the credible intervals.

### Benchmarking the model's descriptive fidelity

Having established the model's internal consistency, we next benchmarked its descriptive performance against conventional, widely adopted analytical methods. The objective of this comparison was not to declare one method universally *superior*, but rather to quantify the fundamental differences in how these disparate analytical frameworks represent and summarize dynamic signals, and to highlight the specific conditions under which our proposed model offers a more advantageous description. The same synthetic datasets from the previous section were analyzed using a 60-second sliding-window analysis for time-domain metrics (mean RRi and SDNN), and a Short-Time Fourier Transform (STFT) with a similar window length for characterizing spectral dynamics.

The results of this comparative analysis underscore the descriptive distortions that are inherent to any method that relies on assumptions of local stationarity when applied to a continuously varying, non-stationary signal. Figure 3 illustrates the outputs from the windowed methods. Visually, these methods produced a delayed and morphologically blunted approximation of the underlying dynamics. The sharp, non-linear transitions that characterize the onset and recovery phases of the signal are smeared and flattened by the averaging process inherent in the windowed calculation. Consequently, the windowed approach failed to accurately capture both the shape and the precise timing of the recovery process. Instead of reflecting the smooth, logistic curve defined by the ground-truth process, the windowed analysis superimposed the square-wave artifacts of its own computational structure onto the data, a clear instance of the analytical method distorting the phenomenon it is intended to measure.

![](figures/fig-windowed-method.svg){width=100%}

**Figure 3. Distorted reconstruction of synthetic data by conventional windowed Methods.** The estimates from a 60-second sliding window (for RRi and SDNN) and an STFT (for spectral proportions) are shown as solid lines overlaid on the ground-truth dynamics (dashed lines). The methods introduce significant lag, blunting, and temporal smearing, failing to capture the true non-linear dynamics.

A formal quantitative analysis confirms the descriptive efficiency and accuracy of the generative approach for signals of this particular nature. Table 3 provides a summary of key performance metrics. Across all scenarios and all signal components, the generative model consistently provided a more accurate and parsimonious description of the data. It achieved substantially lower error (RMSE and Bias) and captured a significantly higher proportion of the signal's variance, with $R^2$ values typically exceeding 0.90. It is important to interpret this result with caution. The model's performance is, in one sense, expected, as the synthetic data were generated from its own structure. However, this does not render the comparison moot. Rather, it precisely illustrates a core theoretical argument of this work: for signals that are plausibly characterized by smooth, continuous transitions between distinct physiological states, a model whose parametric form explicitly encodes this structure will naturally provide a more efficient, less distorted, and more interpretable quantitative summary than a non-parametric, windowed approach that was designed for stationary or quasi-stationary signals.

```{r}
#| column: screen-inset-left
readRDS(file = "tables/tbl-3.RDS")
```

**Table 3. Quantitative Performance Metrics for the Generative Model versus Windowed Methods.** This table presents the Root Mean Squared Error (RMSE), Bias, and Coefficient of Determination ($R^2$) for each method's ability to reconstruct the ground-truth trajectories from all three simulation scenarios. Lower RMSE/Bias and higher $R^2$ indicate superior performance.

## A phenomenological description of an empirical exercise response

To illustrate the framework's intended application in a real-world context, we analyzed an RRi time series recorded from a single healthy subject undergoing a standardized two-minute submaximal step test. The express purpose of this N-of-1 analysis is not to draw generalizable physiological conclusions or to make claims about the nature of exercise physiology. Instead, its function is purely illustrative: to demonstrate how the model translates a complex, high-dimensional, and noisy time series into a concise, low-dimensional, and interpretable set of parameters that form the basis for generating new, quantitative, and ultimately falsifiable hypotheses.

Upon fitting the model to the empirical data, it yields a phenomenological decomposition of the entire exercise-recovery response, as depicted in Figure 4. The framework parses the complex signal into its constituent components, describing the physiological process as a rapid initial tachycardia (a decrease in the baseline RRi component) and a concurrent, profound suppression of total heart rate variability (a decrease in the SDNN component). This initial response is followed by a slower, more gradual recovery phase after the cessation of the exercise stimulus. From a spectral perspective, the model attributes the signature of this response to a near-complete withdrawal of power in the high-frequency (HF) band, a correlate of parasympathetic activity, and a corresponding surge in the proportion of very-low-frequency (VLF) power, a band whose physiological interpretation remains debated but is often associated with sympathetic and other systemic influences. The model captures these dynamics not as a series of disjointed points but as continuous, smooth trajectories, providing a holistic and integrated narrative of the entire physiological event.

![](figures/fig-poc.svg){width=100%}

**Figure 4. Phenomenological decomposition of an empirical exercise-recovery response.** (A) Raw RRi data (black line). (B) The reconstructed RRi signal (red line) overlaid the observed RRi signal (gray). (C-D) time-domain components: the baseline RRi trend ($\mathrm{RR_{base}}$, blue) and the total instantaneous variability (SDNN, orange). (E) The inferred dynamics of the spectral proportions. Shaded regions are 95% highest density intervals.

The framework's primary scientific utility, however, lies not in this qualitative description but in its generation of quantitative, testable claims, which are encapsulated by the posterior distributions of its parameters (Figure 5). A key feature of the model is its separate parameterization of the recovery dynamics of the signal's first moment (mean RRi, via the coefficient $c_r$) and its second moment (total variability, or SDNN, via the coefficient $c_s$). This explicit separation allows for a direct, quantitative description of their potential divergence. In this particular recording, the posterior distributions for these two recovery parameters were credibly distinct; the 95% credible interval for $c_r$ was clearly separated from that for $c_s$. This descriptive feature of the model's output, which we term model-inferred dissonance, provides a precise, data-driven, and falsifiable hypothesis: that the recovery of mean heart rate and the recovery of its variability can follow distinct and dissociable time courses in response to an acute exercise stressor. This hypothesis is not merely a qualitative observation but a quantitative statement, grounded in the posterior probability distributions of the model's parameters.

![](figures/fig-poc-trace.svg){width=100%}

**Figure 5. Traceplots of phenomenological recovery parameters from the empirical analysis.** The stationary distribution for the recovered coefficients of transient changes in mean RRi, total SDNN, and spectral balance, alongside the structured variance fraction ($w$). The traceplots allows for visual inspection of model convergence and potential divergence between the posteriors for model parameters, and provides a quantitative basis for the hypothesis of dissonant autonomic recovery.

# Discussion

The quantitative analysis of HRV has long been a principal method for the non-invasive study of the autonomic nervous system. A persistent and fundamental challenge in this field is the inherent non-stationarity of physiological signals, which directly conflicts with the foundational assumptions of traditional analytical techniques, which were often adopted due to the computational constraints of their era. This discrepancy necessitates the use of methodological compromises, most notably time-windowing, that can introduce significant analytical artifacts and obscure the very dynamics they seek to quantify. The present study was undertaken to develop and evaluate a generative probabilistic model for RRi time series, explicitly designed to represent and infer the parameters of transient physiological events. Our findings, based on a comprehensive suite of simulations and an empirical application, suggest that this inferential paradigm offers a more accurate and granular characterization of autonomic dynamics than is achievable with conventional windowed approaches. This discussion aims to critically interrogate these findings, situate the model within the broader landscape of computational biosignal analysis, conduct an exhaustive examination of its limitations and foundational assumptions, and propose a disciplined path forward for future inquiry.

The primary contribution of this research is not the invention of a fundamentally new class of statistical algorithm, but rather the novel synthesis and application of established principles from modern Bayesian inference to construct a purpose-built, physiologically-oriented measurement instrument. The model formalizes a specific, falsifiable hypothesis about the data-generating process underlying dynamic RRi signals. We employ the term *generative* to signify this complete probabilistic specification, which enables a coherent inferential procedure regarding the model's latent parameters. The internal validity and statistical integrity of this inferential process were interrogated through extensive simulation studies. In these controlled in-silico experiments, where the ground truth was known, the model consistently recovered the generative parameters, and its posterior credible intervals were well-calibrated. This provides baseline confidence that the model can function as a reliable measurement device, but only under conditions where its numerous and restrictive structural assumptions are perfectly met. When benchmarked against conventional methods, its superior performance in tracking dynamic signals was quantitatively demonstrated. This outperformance is an expected consequence of its design, which replaces the crude filtering of windowing with a formal, continuous model of the underlying process. However, the apparent sophistication of this approach belies a series of deep-seated assumptions and conceptual challenges that demand scrupulous consideration.

## Methodological superiority and the abandonment of stationarity

A core assertion of this work is that the analytical shortcomings of traditional HRV analysis are conceptual, not merely technical. Windowed methods are, in essence, digital filters. As such, they inevitably distort the signal they are intended to measure, introducing lag, attenuating amplitudes, and enforcing a false dichotomy between temporal and spectral resolution as dictated by the Heisenberg-Gabor uncertainty principle. The inferential approach adopted here avoids these specific artifacts by modeling the entire time-course of an event holistically. Information from every data point is leveraged to inform the estimate at every other data point, conditioned on the assumed model structure.

However, this approach introduces its own set of profound abstractions. A crucial, and potentially problematic, assumption is the model's imposition of orthogonality on a non-orthogonal system. The model specifies the observed RRi signal as a linear, additive combination of a baseline trend, several sinusoidal components, and a residual noise term. This mathematical decomposition is convenient and ensures parameter identifiability, but it is a gross simplification of cardiovascular physiology. This is akin to analyzing a symphonic recording by perfectly isolating the audio tracks for the violins, cellos, and woodwinds. While one obtains a clean signal for each section, the essential musical information contained within their harmonic interplay and dynamic interaction is irrevocably lost.

Similarly, in physiology, the systems underlying these model components are deeply and non-linearly coupled. Baroreflex activity (a major contributor to the low-frequency component) and respiratory sinus arrhythmia (the high-frequency component) are not independent processes that simply sum together; they are neurologically intertwined within the brainstem and are engaged in a constant, dynamic feedback loop. By construction, the model is forced to assign variance to its clean, orthogonal components in a way that may not map cleanly onto the underlying biology. The estimate for the amplitude of the high-frequency component, for instance, should be interpreted not as a pure measure of respiratory sinus arrhythmia, but as the magnitude of the best-fitting sinusoidal component in that band, assuming it can be linearly and additively separated from all other ongoing dynamics. This is a critical caveat for the physiological interpretation of every parameter yielded by the model.

## From statistical constructs to biological reality

The purported value of this framework is its ability to distill a complex signal into a set of "interpretable" parameters. This claimed interpretability, however, occupies a precarious position between useful abstraction and misleading reification. The risk of reification (i.e., confusing a model's abstract construct for a concrete entity), is acute. The very act of assigning a label like "recovery rate" to a parameter ($\phi$) cognitively biases the researcher to perceive it as a distinct, tangible biological process. While the parameter perfectly describes a feature of the fitted mathematical curve, it is an unsubstantiated inferential leap to equate it with a singular neurophysiological mechanism. The observed recovery trajectory is the net result of an immensely complex cascade of interacting neural, endocrine, and humoral processes. The model's elegant parameterization of this trajectory is a descriptive summary, not an elucidation of its constituent parts.

This interpretive discipline must be rigorously applied to all of the model's outputs. The concept of "dissonant autonomic recovery", for example, is a statistically identifiable pattern where heart rate returns to baseline far more quickly than measures of heart rate variability. Yet, its proposed link to "lingering sympathetic activation" remains a speculative hypothesis requiring external validation. Could this pattern alternatively reflect differential downregulation rates of adrenergic and cholinergic receptors, or a mismatch in the resetting of central autonomic command versus peripheral feedback? These are precisely the kinds of sophisticated hypotheses the model allows one to formulate, but it possesses no inherent capacity to adjudicate between them. It provides the tool to quantify the phenomenon with precision, thereby rendering such hypotheses testable, but it cannot, on its own, provide the causal evidence.

Nowhere is this interpretive caution more warranted than with the structured variance fraction, $w$. Its proposed role as an index of "autonomic coherence" is a compelling narrative, but one that rests on a fragile evidentiary foundation. At its core, $w$ is simply the proportion of signal variance captured by a fixed set of sinusoids. A reduction in $w$ could plausibly reflect a pathological dysregulation of autonomic rhythms. It could equally reflect a host of mundane confounders: an increase in movement artifact, the presence of uncorrected ectopic beats, a change in breathing frequency outside the canonical high-frequency band, or activation of other physiological processes that contribute to RRi variance but are not sinusoidal. To label this parameter with a term as specific as "autonomic coherence" is premature and risks imbuing it with a degree of physiological certainty it does not currently warrant. It is more accurately described as an "index of rhythmic regularity", a term that is less evocative but more intellectually honest.

## Limitations

The credibility of this work demands a comprehensive inventory of its limitations, which span from the foundational to the practical.

The model's architecture is built upon a series of rigid and potentially invalid assumptions. The choice of a symmetric double-logistic function is a matter of mathematical convenience, not physiological imperative; asymmetric biological recovery processes are common and would be poorly approximated by this functional form. More fundamentally, the model is a specialist instrument designed for a single, isolated perturbation-recovery event. This renders it highly effective for analyzing clean, stimulus-response experimental paradigms but makes it fundamentally unsuited for the exploratory analysis of complex, naturalistic data containing multiple, overlapping events of unknown form. It is a tool for confirmatory, not exploratory, science.

The adoption of a Bayesian framework, while providing intuitive probabilistic outputs, comes with its own philosophical baggage. The posterior credible interval for a parameter is a statement of subjective belief about its value, conditional on the data, the chosen model structure, and the specified prior distributions. It is not equivalent to a frequentist confidence interval and does not account for structural uncertainty, the profound possibility that the model itself is misspecified. This leads to what might be termed the "Rashomon effect" in modeling, where multiple, structurally different models can provide equally good fits to the data while telling fundamentally different and contradictory stories about the underlying process. Without a systematic exploration of this "model space", the conclusions drawn from this single, handcrafted model must be regarded as highly provisional.

The model's reliance on Hamiltonian Monte Carlo sampling makes it computationally expensive. The required runtime may not only be a practical barrier for large datasets but could also create an "analytical divide", where such advanced methods are only accessible to research groups with substantial computational resources. This effectively sidelines the method from deployment in real-time or near-real-time clinical settings, such as intensive care monitoring, where rapid feedback is paramount. Furthermore, the model's performance is inextricably linked to upstream data fidelity. Its sophisticated inference engine is being applied to RRi data that has already undergone pre-processing to remove artifacts. Uncorrected premature ventricular contractions, for instance, can drastically alter the local mean and variance. While the model may be robust to small deviations, its behavior in the face of such gross, non-stationary artifacts is unknown and could lead to biased inference on the entire recovery curve. Finally, the treatment of the residual term as unstructured "noise" is a convenient fiction. In biology, true randomness is rare; the residuals likely contain unmodeled, structured biological processes or deterministic chaos. The assumption of a simple, stationary Gaussian noise process is a strong one that requires, but has not yet received, rigorous justification through posterior predictive checks and deep residual analysis.

## Future directions

The extensive limitations catalogued above do not invalidate the approach but rather define a rigorous and necessary line for future research. This line of research should move beyond incremental model refinement and toward the systematic characterization of the model as a scientific instrument.

The most urgent task is to begin the process of instrument calibration. Just as a new telescope must have its optical aberrations mapped and its resolution quantified, this model requires a comprehensive characterization of its operational properties. An ideal future study would involve establishing a formal calibration protocol: creating a standardized open-source library of synthetic RRi signals with known complexities, asymmetric responses, overlapping events, various artifact types, and non-Gaussian noise, and then systematically testing the model against this library. The results, published as a detailed "specification sheet", would delineate the model's biases, sensitivities, and failure modes, defining its safe operating envelope for future users.

Furthermore, future work must focus on using the model for quantitative hypothesis falsification. The framework should be applied in experimental contexts where specific interventions are known to target certain physiological pathways. For instance, testing the effect of beta-blockade on the response amplitude and recovery rate parameters following a cold pressor test would provide crucial external validation, linking the model's abstract parameters to tangible pharmacological effects. This shifts the focus from "fitting a model" to "using a model to test a theory".

In conclusion, the generative framework presented here represents a conceptually powerful alternative to conventional methods for analyzing dynamic HRV. Its true contribution, however, must be framed with profound intellectual humility. This model is not a window into reality; it is an epistemic instrument, a purpose-built lens for viewing complex physiological data through a particular, highly structured, and simplified mathematical frame. Its value lies not in its verisimilitude, but in its ability to translate a noisy, high-dimensional signal into a sparse set of quantitative parameters about which precise, falsifiable scientific arguments can be constructed. The difficult, painstaking work of calibrating this instrument, understanding its distortions, and cautiously deploying it to test concrete physiological hypotheses will ultimately determine its lasting contribution to the science of autonomic control.

# Methods

## Model Formulation

We presented a comprehensive generative model for the RRi signal, observed at a discrete set of time points $\{t_i\}_{i=1}^N$. The model's architecture is designed to deconstruct the signal into its constituent physiological components, providing a mechanistic account of the processes that shape heart rate dynamics. This is achieved by conceptualizing the signal as a probabilistic process controlled by a time-varying mean, $\mu(t_i)$, and a partitioned, time-varying standard deviation. This sophisticated approach moves beyond simple curve-fitting to formalize a theory of autonomic control and its temporal evolution within a fully Bayesian framework.

The cornerstone of the model is the partitioning of the total signal variance into two distinct, time-dependent components: a structured variance, $\sigma^2_{\text{struct}}(t_i)$, which captures the physiologically meaningful, oscillatory patterns of HRV, and a residual variance, $\sigma^2_{\text{resid}}(t_i)$, which accounts for unstructured, moment-to-moment fluctuations or measurement noise. The complete observation model, which integrates these components, is defined by the Normal likelihood in @eq-full-likelihood.

$$
\mathrm{RRi}(t_i) \sim \mathcal{N}\left(\mu(t_i), \sigma_{\text{resid}}(t_i)\right)
$${#eq-full-likelihood}

The core of the model lies in the construction of the mean and variance components from a shared set of underlying dynamic functions. The mean trajectory, $\mu(t_i)$, is a superposition of a smoothly varying baseline trend and the synthesized structured signal itself, as shown in @eq-mean-model.

$$
\mu(t_i)
= \underbrace{\mathrm{RR_{base}}(t_i)}_{\substack{\text{Gross}\\\text{RRi Trend}}}
+
\underbrace{A(t_i) \cdot \sum_{j=1}^{J} p_j(t_i) \cdot S_j(t_i)}_{\substack{\text{Structured Variability Signal}}}
$${#eq-mean-model}

Here, $\mathrm{RR_{base}}(t_i)$ represents the gross, underlying heart period trajectory. The structured variability signal is synthesized from a set of dynamically evolving spectral oscillators, $S_j(t_i)$, which represent activity in different physiological frequency bands ($j=1,2,3$ for VLF, LF, and HF, respectively). These oscillators are weighted by time-varying proportions, $p_j(t_i)$, and their overall magnitude is controlled by a scaling amplitude, $A(t_i)$. This amplitude is deterministically calculated to ensure the signal's variance exactly matches the target structured variance, $\sigma^2_{\text{struct}}(t_i)$.

A pivotal feature is a dynamic trajectory for the total signal variability, denoted $\mathrm{SDNN}(t_i)$, which is then partitioned into its structured and residual components. This partition is controlled by a parameter, $w \in [0, 1]$, which represents the fraction of total variance attributable to the structured component, as defined in @eq-var-structure.

$$
\begin{aligned}
\sigma^2_{\text{struct}}(t_i) &= w \cdot \mathrm{SDNN}(t_i)^2 \\
\sigma^2_{\text{resid}}(t_i) &= (1-w) \cdot \mathrm{SDNN}(t_i)^2
\end{aligned}
$${#eq-var-structure}

This formulation allows the model to simultaneously learn not only how the total amount of HRV changes over time (via $\mathrm{SDNN}(t_i)$), but also how the nature of that variability evolves, that is the balance between predictable, oscillatory patterns and unpredictable noise (via $w$). This unified analysis facilitates a deeper understanding of autonomic regulation by capturing phenomena across both time and frequency domains within a single, coherent inferential framework.

### Baseline Heart Period and Total Variability Trajectories

The model posits that the primary trends in both the mean heart period and its total variability are driven by a common underlying physiological response to a stimulus. To capture this, both the $\mathrm{RR_{base}}(t_i)$ and $\mathrm{SDNN}(t_i)$ trajectories are parameterized using the same flexible double-logistic functional form.

The baseline heart period, $\mathrm{RR_{base}}(t_i)$, which quantifies the gross, underlying variations in the mean R–R interval, is defined in @eq-rri-baseline-model.

$$
\mathrm{RR_{base}}(t_i) =
\underbrace{\alpha_r}_{\substack{\text{Resting RRi}}}
-
\underbrace{\beta_r \cdot \mathcal{D}_{1}(t_i)}_{\substack{\text{Perturbation-induced}\\\text{RRi Drop}}}
+
\underbrace{c_r \beta_r \cdot \mathcal{D}_{2}(t_i)}_{\substack{\text{Post-perturbation}\\\text{RRi Recovery}}}
$${#eq-rri-baseline-model}

In this formulation, $\alpha_r$ represents the initial, stable heart period. The parameter $\beta_r$ signifies the magnitude of the decline in RRi induced by the perturbation. The fractional recovery amplitude is denoted by $c_r$, where $c_r > 1$ indicates an overshoot.

Concurrently, the total instantaneous variability, $\mathrm{SDNN}(t_i)$, follows a parallel dynamic trajectory, as defined in @eq-sdnn-model.

$$
\mathrm{SDNN}(t_i) =
\underbrace{\alpha_s}_{\substack{\text{Resting SDNN}}}
-
\underbrace{\beta_s \cdot \mathcal{D}_{1}(t_i)}_{\substack{\text{Perturbation-induced}\\\text{SDNN Drop}}}
+
\underbrace{c_s \beta_s \cdot \mathcal{D}_{2}(t_i)}_{\substack{\text{Post-perturbation}\\\text{SDNN Recovery}}}
$${#eq-sdnn-model}

Here, $\alpha_s$, $\beta_s$, and $c_s$ are analogous to their counterparts in the baseline model, representing the resting total SDNN, the magnitude of its suppression, and its fractional recovery, respectively. The dynamics for both trajectories are driven by a shared pair of logistic transition functions, $\mathcal{D}_{1}(t_i)$ and $\mathcal{D}_{2}(t_i)$, defined in @eq-logistic-components.

$$
\begin{aligned}
\mathcal{D}_{1}(t_i) &= \left(1+ e^{-\lambda (t_i - \tau)}\right)^{-1} \\
\mathcal{D}_{2}(t_i) &= \left(1+ e^{-\phi (t_i - \tau - \delta)}\right)^{-1}
\end{aligned}
$${#eq-logistic-components}

The shared timing parameters enforce a strong physiological coupling: $\tau$ is the inflection point of the initial response, with rate $\lambda$. The second transition, representing recovery, is offset by a delay $\delta$ and proceeds at a rate $\phi$. This shared structure ensures that changes in the magnitude of variability are temporally synchronized with changes in the mean heart period, while still allowing their relative magnitudes and recovery profiles to differ.

### Generative Model for the Structured Signal

The structured component of the signal is where the model's spectral properties are defined. Its construction involves three key elements: the dynamic spectral proportions, the latent spectral oscillators (whose amplitudes are governed by a Gaussian Process), and a deterministic amplitude inversion that ties them to the target variance.

#### Dynamic Frequency Band Proportions

The proportions $p_j(t_i)$ dictate how the total structured variance, $\sigma^2_{\text{struct}}(t_i)$, is allocated across the different frequency bands at each moment. The model captures the evolution of these proportions as a smooth transition between two distinct spectral states: a baseline state ($\vec\pi_{\text{base}}$) and a perturbed state ($\vec\pi_{\text{pert}}$). The transition is determined by a single master controller function, $C(t_i)$, as shown in the convex combination of @eq-pj-using-ct.

$$
\vec{p}(t_i) = (1 - C(t_i)) \cdot \vec\pi_{\text{base}} + C(t_i) \cdot \vec\pi_{\text{pert}}
$${#eq-pj-using-ct}

Here, $\vec\pi_{\text{base}}$ and $\vec\pi_{\text{pert}}$ are simplex vectors representing the characteristic spectral distributions at rest and during peak perturbation. The master controller, $C(t_i)$, is itself built from the same logistic building blocks, ensuring the spectral transition is synchronized with the primary physiological response, as defined in @eq-master-controller.

$$
C(t_i) = \mathcal{D}_{1}(t_i) \cdot \left(1 - c_c \cdot \mathcal{D}_{2}(t_i)\right)
$${#eq-master-controller}

This function naturally transitions from 0 towards 1, with the parameter $c_c \in [0, 1]$ allowing for an incomplete spectral recovery.

#### Latent Spectral Oscillators via a Gaussian Process Prior

A central task of the model is to infer the power of oscillations across a range of frequencies from the observed data. A simple approach might impose a rigid mathematical structure on the spectrum (e.g., a $1/f^b$ power law) or treat the power at each frequency as independent. However, physiological spectra are rarely so simple; they typically exhibit smooth, continuous shapes with broad peaks.

To capture this realistic structure, we employ a flexible, non-parametric Gaussian Process (GP) prior. A GP is a statistical tool that allows us to place a prior on an unknown function, enforcing the assumption that it is smooth. In this context, it formalizes the intuition that the amplitudes of nearby frequencies should be similar, allowing the model to learn the characteristic shape of the spectrum within each band directly from the data.

Each underlying oscillator, $S_j(t_i)$, which contributes to the structured variability, is modeled as a sum of simple sine and cosine waves with pre-specified frequencies ($f_{j,k}$) but unknown amplitudes ($u_{j,k}$), as shown in @eq-multi-sine-fun.

$$
S_j(t_i) = \sum_{k=1}^{K_j} \left[ u_{j,k}^{(\sin)} \sin(2 \pi f_{j,k} t_i) + u_{j,k}^{(\cos)} \cos(2 \pi f_{j,k} t_i) \right]
$${#eq-multi-sine-fun}

The key innovation is the prior placed on these amplitudes. This is a multi-step process designed for both flexibility and computational stability. First, we model the underlying log-amplitude envelope, a vector denoted $\vec{v}_j$, as a single draw from a GP. The GP is defined by a zero-mean function and a squared exponential covariance kernel ($\mathbf{K}_j$), which determines the smoothness of the spectral shape, defined in @eq-gp-kernel.

$$
\vec{v}_j \sim \mathcal{GP}(\vec{0}, \mathbf{K}_j), \quad \text{where } (\mathbf{K}_j)_{kl} = \alpha_{gp,j}^2 \exp\left(-\frac{(f_{j,k} - f_{j,l})^2}{2\rho_{gp,j}^2}\right)
$${#eq-gp-kernel}

The covariance matrix $\mathbf{K}_j$ is governed by two hyperparameters which are themselves estimated from the data: $\alpha_{gp,j}$, which correspond to the marginal standard deviation, which controls the overall magnitude or "waviness" of the spectral shape. A larger $\alpha_{gp,j}$ allows for larger peaks and deeper troughs in the spectrum. Moreover, $\rho_{gp,j}$ is the length-scale, which controls the smoothness of the function. A large $\rho_{gp,j}$ enforces a very smooth, slowly changing shape, while a small value allows for more rapid variations and narrower spectral peaks.

To ensure this complex model can be estimated efficiently, we use a non-centered parameterization. This is a computational technique that improves the performance of the sampler by reducing correlations between parameters. Instead of estimating the highly correlated vector $\vec{v}_j$ directly, the model estimates a vector of independent standard normal deviates, $\vec{z}_{gp,j}$. The target log-amplitude envelope is then deterministically constructed via the transformation $\vec{v}_j = \mathbf{L}_j \vec{z}_{gp,j}$, where $\mathbf{L}_j$ is the Cholesky factor of the covariance matrix ($\mathbf{K}_j = \mathbf{L}_j \mathbf{L}_j^\top$). This approach makes the estimation problem much simpler for the algorithm, leading to more robust and reliable inference.

#### Normalization for Identifiability

A critical step in the model's construction is to ensure that its parameters are identifiable, meaning there is a unique set of parameters that explains the data. A potential ambiguity arises because two components of the model control amplitude: the GP's marginal standard deviation ($\alpha_{gp,j}$) and the overall signal amplitude, $A(t_i)$. Without a constraint, the model could achieve the same result by increasing the GP's amplitude while decreasing $A(t_i)$, making it difficult to interpret either parameter.

To resolve this, we enforce a clear division of labor. The Gaussian Process is assigned the exclusive role of determining the relative shape of the spectrum, while $A(t_i)$ is assigned the role of setting the absolute power.

This is achieved by normalizing the amplitude envelope derived from the GP. First, let $\vec{a}_j = \exp(\vec{v}_j)$ be the vector of positive amplitudes determined by the GP. We then create a normalized scaling vector, $\vec{s}_j$, such that the expected variance of the synthesized oscillator $S_j(t_i)$ is precisely 1.

Because the final coefficients $u_{j,k}$ are generated by scaling standard normal deviates ($z_{j,k} \sim \mathcal{N}(0,1)$), the appropriate normalization must be based on the expected variance of this random process. A key property of this process is that the expected value of cross-products between independent random coefficients is zero. Consequently, the total expected variance simplifies to a sum that depends only on the diagonal elements of the pre-computed Gram matrices ($\mathbf{G}_{\sin,j}$ and $\mathbf{G}_{\cos,j}$). This procedure is formalized in @eq-normalization.

$$
\vec{s}_j = \frac{\vec{a}_j}{\sqrt{\sum_{k=1}^{K_j} a_{j,k}^2 \left( (\mathbf{G}_{\sin,j})_{kk} + (\mathbf{G}_{\cos,j})_{kk} \right)}}
$${#eq-normalization}

The final oscillator coefficients are then constructed as $u_{j,k}^{(\sin)} = z_{j,k}^{(\sin)} \cdot s_{j,k}$ and $u_{j,k}^{(\cos)} = z_{j,k}^{(\cos)} \cdot s_{j,k}$. This procedure ensures that the GP component only informs the spectral shape, normalized to have a unit expected variance, thereby removing the ambiguity and making the model's parameters clearly interpretable.

#### Deterministic Amplitude Inversion

The final step is to ensure that the synthesized structured signal, $X(t_i) = A(t_i) \sum_{j=1}^J p_j(t_i) S_j(t_i)$, has a variance equal to the target structured variance, $\sigma^2_{\text{struct}}(t_i)$, at every time point. A key feature of this model is that it does not assume the sinusoids form an orthogonal basis over the discrete, finite time domain. Instead, it computes the exact sample variance for each synthesized oscillator, $\mathrm{Var}[S_j(t_i)]$, given the current values of its coefficients.

Let $\mathbf{\Sigma}_S$ be the diagonal $3 \times 3$ matrix whose entries are these computed variances, $\mathbf{\Sigma}_S[j, j] = \mathrm{Var}[S_j(t_i)]$. Assuming independence between the bands, the variance of the weighted sum is given by the quadratic form in @eq-weighted-sum.

$$
\mathrm{Var}\left[\sum_{j=1}^J p_j(t_i) S_j(t_i)\right] = \vec{p}(t_i)^\top \mathbf{\Sigma}_S \; \vec{p}(t_i)
$${#eq-weighted-sum}

The variance of the complete structured signal is $\mathrm{Var}[X(t_i)] = A(t_i)^2 \cdot \left(\vec{p}(t_i)^\top \mathbf{\Sigma}_S \; \vec{p}(t_i)\right)$. To match our target, we set this equal to $\sigma^2_{\text{struct}}(t_i) = w \cdot \mathrm{SDNN}(t_i)^2$. Solving for $A(t_i)$ yields the critical inversion formula in @eq-amplitude-inversion.

$$
A(t_i) = \frac{\sqrt{w} \cdot \mathrm{SDNN}(t_i)}{\sqrt{\vec{p}(t_i)^\top \mathbf{\Sigma}_S \; \vec{p}(t_i)}}
$${#eq-amplitude-inversion}

This inversion is essential. It dynamically adjusts the amplitude of the synthesized spectral signal to ensure its contribution to the total variance is exactly as prescribed by the model's high-level parameters ($\mathrm{SDNN}(t_i)$ and $w$), thereby connecting the time-domain and frequency-domain components of the model.

### Model Parameterization and Priors

To ensure numerical stability and efficient sampling, all model parameters are estimated on an unconstrained real-valued scale ($\mathbb{R}$). Priors are placed on these unconstrained parameters, which are then transformed back to their constrained, physically meaningful scales within the model.

#### Timing, Rate, and Magnitude Parameters

Parameters constrained to a specific interval are parameterized on the logit scale, while positive parameters are on the log scale. For instance, the timing parameters $\tau$ and $\delta$ are mapped to the observed time interval ($t_\text{range}$) and the remaining time, respectively, using the inverse logit transformation, such that $\tau = \text{logit}^{-1}(\tau_\text{logit}) \cdot t_\text{range} + t_\text{min}$ and $\delta = \text{logit}^{-1}(\delta_\text{logit}) \cdot (t_\text{range} - \tau)$.

The positive rate parameters $\lambda$ and $\phi$ are simply log-transformed, i.e., $\lambda = \exp(\lambda_\text{log})$. The recovery coefficients for the time-domain components ($c_r,c_s$) are mapped to the interval $[0, 2]$, while the recovery parameter for spectral components ($c_c$) is mapped to the $[0, 1]$ interval, using a scaled logit function. Similarly, the fraction of structured variance, $w$, is constrained between 0 and 1 via $w = \text{logit}^{-1}(w_{\text{logit}})$.

The magnitude parameters ($\alpha_r, \beta_r, \alpha_s, \beta_s$) are scaled relative to data-derived quantities to create dimensionless parameters whose priors are easier to specify. These transformations ensure the sampler explores a valid and well-behaved parameter space.

#### Spectral and Oscillator Parameters

The spectral proportions, $\vec\pi_{\text{base}}$ and $\vec\pi_{\text{pert}}$, are mapped from the 3-dimensional simplex to 2-dimensional real vectors using the additive log-ratio (ALR) transformation. The model estimates the unconstrained vectors, which are mapped back to the simplex via the softmax function.

The GP hyperparameters are given weakly informative, regularizing priors. The marginal standard deviation is given a half-Normal prior, $\alpha_{gp} \sim \mathcal{N}^+(0, 0.5)$, favoring modest spectral amplitudes. The length-scale is given a $\text{Log-normal}(0, 0.5)$ prior, regularizing the smoothness of the spectral shape.

Finally, the non-centered parameterization is completed by placing standard Normal priors on all latent "z" variables: the GP deviates ($z_{gp}$), the sine coefficients ($z_{\sin}$), and the cosine coefficients ($z_{\cos}$). This technique is critical for efficient sampling, as it decouples the hierarchical dependencies and removes pathological posterior geometries.

## Simulation Studies

To evaluate the model's performance, we conducted a series of simulation studies. The objectives were to assess the model's ability to recover known ground-truth parameters from synthetic data and to quantitatively benchmark its reconstruction of key HRV dynamics against conventional, windowed analysis techniques. Synthetic RRi time series were generated using the model's own generative structure with pre-specified parameter values. This process provided datasets with perfectly known underlying dynamics for $\mathrm{RR}(t_i)$, $\mathrm{SDNN}(t_i)$, and $p_j(t_i)$, serving as an objective gold standard for evaluation.

We designed three distinct scenarios to probe the model's capabilities under diverse and challenging conditions. These scenarios: (A) a classic sympatho-vagal response, (B) an incomplete recovery with time-domain and spectral persistence, and (C) an incomplete spectral but complete time-domain recovery with high noise, were chosen specifically to test the limitations inherent in standard analysis methods. Each simulated dataset was analyzed with our full Bayesian model and, for comparison, with conventional techniques: a sliding-window analysis for time-domain metrics and a Short-Time Fourier Transform (STFT) for spectral dynamics.

### Time-domain trajectory reconstruction

The capacity to accurately capture time-domain dynamics was evaluated by comparing the model's continuous estimates of $\mathrm{RR}(t_i)$ and $\mathrm{SDNN}(t_i)$ against the stepwise approximations derived from a standard sliding-window mean and standard deviation. To provide a comprehensive evaluation of reconstruction fidelity, model performance was quantified against the ground-truth trajectories using a suite of complementary metrics. The root mean squared error (RMSE) and mean absolute error (MAE) were used to measure the average magnitude of the estimation error, with MAE offering a more robust assessment against sporadic outliers. The model's bias (mean error) was calculated to identify any systematic tendency for over- or underestimation. Finally, the coefficient of determination ($R^2$) was used to assess the proportion of variance in the true signal captured by the estimate.

### Spectral dynamics reconstruction

To evaluate the reconstruction of spectral dynamics, our model's continuous estimates of the spectral proportions, $p_j(t_i)$, were compared against those derived from an STFT. For a direct comparison, the spectrogram generated by the STFT was normalized at each time step to yield proportional power within the VLF, LF, and HF bands. The performance of both methods was then assessed by comparing their estimated trajectories to the known ground-truth proportions using the same comprehensive suite of metrics (RMSE, MAE, Bias, and $R^2$). This comparative framework was designed to objectively quantify the trade-offs in temporal and frequency resolution inherent to windowed methods versus the continuous estimation provided by our model, especially in the presence of rapid transitions or low signal-to-noise ratios. As part of the model's internal validation, we also confirmed that the 95% posterior credible intervals for all generative parameters consistently contained the known ground-truth values, ensuring both inferential accuracy and appropriate uncertainty quantification.


# References

<div id="refs"></div>

# Author Contributions

Conceptualization, MC-A, CN-E; Data curation, MC-A; Investigation, MC-A; Methodology, MC-A; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E; Writing–review & editing, MC-A, [...], CN-E. All authors have read and agreed to the published version of the manuscript.

# Funding

This work was funded by ANID Proyecto Fondecyt Regular Nº1250474 and by the Innovation Fund for Competitiveness of the Regional Government of Magallanes and Chilean Antarctica (BIP Code 40042452-0).

# Data availability statement

The raw data supporting the conclusions of this article will be made available by the authors without undue reservation.

# Conflicts of interests

The authors declare that the research was conducted without any commercial or financial relationships construed as as a potential conflict of interest.
